---
title: "Map of unemployment rates in EU countries, 2022"
author: "Shefali C."
date: "2023-09-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-emo, message=FALSE, warning=FALSE, include=FALSE}
library(emo)
```


This notebook is an attempt to recreate map of member states of European Union as seen in [this](https://ec.europa.eu/eurostat/statistics-explained/index.php?title=Unemployment_statistics_at_regional_level#Regional_variations_in_youth_unemployment) article by Eurostat, using R and **`giscoR`** package.  
I've tried explaining each step in the process; I hope this will be helpful for those getting started with **`giscoR`** and **`eurostat`** packages. `r emo::ji("smile")`  

(**Sidenote:** If you wish to include emojis in your notebook, install the  **`emo`** package.  
And use following inline code to insert emoji: **`\r emo::ji("happy")`**.  
For more details on **emo**, refer [this Github page](https://github.com/hadley/emo).)  


### Introduction:  

#### 1. Process: 

The process to be followed is pretty straight-forward.  
- First we work on the data to be represented in the map, here unemployment rate. This data is fetched using **eurostat** package functions.  
- Then, we fetch the shapefiles for all NUTS-2 regions, countries and their respective boundaries using **giscoR** package.  
- Then, we merge these two dataframes to get one consolidated df with spatial data as well as the content to be displayed.  
- Then we use **`geom_sf()`** function to build the map. **"sf"** means simple features; it refers to the geometrical objects present in our data like polygons, lines or points.

#### 2. Packages used:  

- **`tidyverse`-** Main package for data manipulation and visualizations. I've loaded the entire metapackage as a force of habit, but loading **`dplyr`** and **`ggplot2`** will do the job.  

- **`giscoR`-** This package is needed to get shapefiles for all countries and NUTS-2 regions. Shapefiles here include ***MULTIPOLYGONS*** for shape of each region and ***MULTILINESTRING*** for creating boundaries.  

- **`eurostat`-** Unemployment rate in EU member states for the year 2022 has been fetched using this package.  

- **`sf`-** This package is the backbone for working on spatial data in R. It is required to create maps.  

- **`colorspace`-** While creating the maps, I tried several color palettes for the cholopleth map. **colorspace** helps to visualize color swatches of all the palettes.  

- **`lubridate`-** Used for manipulation of date column in the dataset.  


```{r load-libraries, warning=FALSE, message=FALSE}
#main package for data manipulation and viz
library(tidyverse)
#to fetch datasets
library(giscoR)
library(eurostat)
#sf required for spatial data mapping
library(sf)
#to see color-swatches of palettes
library(colorspace)
#to work on dates column
library(lubridate)
#library(plotly)
```


```{r get-dirs}
working_dir <- getwd()
cache_directory <- file.path(working_dir, "data/eu-unemployment")

```  

- The datasets are of large size. Setting a cache directory here and **`cache = TRUE`** in upcoming data extraction functions helps to download the data in the target directory.  
- The files get stored in **.rds** format and can be read using **`read_rds()`** of `readr` package.  


```{r set-cache-dir, message=FALSE, warning=FALSE}
#set path to cache the dataset
set_eurostat_cache_dir(cache_dir = cache_directory, install = T,
                       overwrite = T)
```  

#### 3. Dataset:  

I've mapped the unemployment rates in EU at NUTS-2 level. Each of these Eurostat datasets have been given a code in the database. The code for data used here is- **lfst_r_lfu3rt**.  

- List of entire datasets available- [Eurostat Database](https://ec.europa.eu/eurostat/data/database).  

- Expand the navigation tree to fetch data of your choice. For labour-force datasets, follow the directions below:  


**Detailed Datasets-->General and regional statistics-->Regional Statistics by NUTS classification (reg)-->Regional labour market statistics (reg_lmk)-->Regional unemployment LFS annual series-->Unemployment rates by sex, age, educational attainment level and NUTS 2 regions (%) (lfst_r_lfu3rt).**  

- To download the data for the first time, use the **`get_eurostat()`** function of `eurostat` package. To download the data in cache directory, keep `cache=T`; to get full labels in each columns instead of code, keep `type = 'labels'`. Default is `type = 'code'`.

For .e.g.,  
 - `type='labels'` gives full statements like "Greater than 25 years", "From 15 to 74 years" in the "age" column.  
 - `type='code'` gives short codes like "Y_GE25" & "Y15-74" for the same values.  
 

```{r data-download}
#data_with_code <- eurostat::get_eurostat(id = "lfst_r_lfu3rt", cache = T)
```

Since, I already have the data downloaded, I've used **`read_rds()`** of `readr` package to read the rds file.  
The baseR function for the same is **`readRDS()`** if you do not wish to use readr.  

```{r read-rds-file}
#read the data_with_code from cache directory
data_with_code <- read_rds(file = paste0(cache_directory, "/lfst_r_lfu3rt_code.rds"))
#make a copy
data_with_code2 <- data_with_code
```  

#### 4. Column description:  

- **isced11-** The education level as defined in the Eurostat database. The full form is ***International Standard Classification of Education (ISCED 2011)***. For e.g.  
      - **ED0-2** refers to less than primary education.  
      - **NRP** refers to No response received.  
      - **TOTAL** means All ISCED 2011 levels.  
      - **UNK** means Unknown.  
      To search what any given code means in this column, use [this data dictionary](https://dd.eionet.europa.eu/vocabulary/eurostat/isced11/view) of Eurostat.  
      
- **age-** contains different age ranges like Y15-24, Y35-44, Y_GE15 etc. I will be using rows with age group Y15-74.  

- **unit-** is nothing but **"PC"**, the measurement unit for unemployment in this dataframe. Here, it is percentage. There is another table with total count of unemployed people, not percentage. It has a different Eurostat code and can be found on the database page linked above.  

- **geo-** contains code of all NUTS levels- 0, 1 and 2 for each country. Note, although, on the database page, this file indicates data for NUTS-2 regions, it contains rows with NUTS-0 and NUTS-1 levels too.  
    For e.g., in case of Austria, unemployment rate is present for:  
    - **"AT"** (Austria), which is the country's estimate,  
    - **"AT1", "AT2", "AT3"** which are NUTS-1 regions- ***Eastern Austria (Ostösterreich), Southern Austria (Südösterreich), Western Austria (Westösterreich)***.  
    - **"AT11", "AT12", "AT13", "AT21", "AT22", "AT31", "AT32", "AT33", "AT34"** which are NUTS-2 level classification.  

- **time-** contains dates ranging from 1999 to 2022. Dates are in format- YYYY-01-01.  

- **values-** contains estimated percentage unemployment rate.  


```{r view-column-names}
colnames(data_with_code2)
```



```{r summary-time-column}
#let's see the time range of this data
summary(data_with_code2$time)
```  

Time column has been converted to type **POSIXlt** in order to apply **`year()`** function of `lubridate` to filter out data for 2022 later.

```{r convert-time}
#convert time column to date format
data_with_code2$time <- as.POSIXlt(data_with_code$time)

```

The data we are concerned with here is unemployment rate in EU at NUTS-2 level for all people (both male and female; no transgender category in sex column.) and all levels of education (denoted by "TOTAL" in the isced11 column.), within the age group 15-74 years.  

Keeping this in mind, some filters have been applied below.  

```{r data-2022}
#filter data for the year 2022
data_2022 <- data_with_code2 %>% filter(year(time) == 2022,
                                        isced11 == "TOTAL", 
                                        sex == "T",
                                        age == "Y15-74")
#remove regions with listed names
data_2022 <- data_2022 %>% filter(!geo %in% c("EU27_2020", "EA20"))
```

In the code below, the unemployment rate % has been divided into 5 intervals: <3.1%, 3.1%-4.5%, 4.5%-6.2%, 6.2%-9.0% and >9.0%.  
The `legend_labels` list contains values that will represent each of these class intervals in legend keys.  

Later on, a color palette has been created with 5 colours. Each of the colors in the choropleth map represent the category in which individual NUTS-2 regions fall in, depending on the unemployment rate for that area in 2022.  

```{r class-intervals}
#create breaks
breaks <- c(0, 3.1, 4.5, 6.2, 9.0, ceiling(max(data_2022$values, na.rm = T))+1)

#create intervals
data_2022$perc_range <- cut(data_2022$values, breaks = breaks,
                            include.lowest = T)

#minor alterations in breaks to make legend labels better
#legend_labels <- prettyNum(breaks)
legend_labels <- c("< 3.1", "3.1 - <4.5", "4.5 - <6.2", "6.2 - <9.0", 
                   ">=9.0", "data not available")
```  

Right now, our dataframe looks something like this:  

```{r df-head}
head(data_2022, 3)
```

Except for columns `geo`, `values` and `perc_range`, rest of the columns same value in all rows, so I'm removing all these irrelevant columns now.  

```{r final-data}
final_data <- data_2022 %>% select(geo, values, perc_range)
```

This dataframe now looks like this:  

```{r final-data-head}
final_data[sample(nrow(final_data),4),]
```

We need to add columns for country name, country code, geometry of each region.  

In the code below, I've explicitly created a dataframe using columns of gisco_countries, in order to prevent the geometry column from being included as it gets tagged along if gisco_countries is directly used.

```{r}
#get names of all countries with their codes
cntr_id <- gisco_countries$CNTR_ID
cntr_name <- gisco_countries$NAME_ENGL
country_codes_names <- data.frame(CNTR_CODE = cntr_id, CNTR_NAME = cntr_name)
```


```{r}
#fetch spatial data for all NUTS 0,1,2 regions

nuts0 <- gisco_get_nuts(year = "2021", cache = T, nuts_level = 0,
                        resolution = "20")
nuts1 <- gisco_get_nuts(year = "2021", nuts_level = 1, resolution = "20",
                        cache = T)
nuts2 <- gisco_get_nuts(year = "2021", cache = T, resolution = "20",
                        nuts_level = 2)

#create a subset of the NUTS-2 df
nuts0_crop <- nuts0 %>% select(NUTS_ID,LEVL_CODE,CNTR_CODE,NAME_LATN, geometry)
nuts1_crop <- nuts1 %>% select(NUTS_ID,LEVL_CODE,CNTR_CODE,NAME_LATN, geometry)
nuts2_crop <- nuts2 %>% select(NUTS_ID, LEVL_CODE,CNTR_CODE,NAME_LATN, geometry)

all_nuts <- dplyr::bind_rows(nuts0_crop, nuts1_crop, nuts2_crop) %>% arrange(NUTS_ID)


data_nuts_merged <- data_2022 %>%
                      select(age, geo, values) %>% 
                      left_join(all_nuts,
                                by = c("geo" = "NUTS_ID"))

```

```{r}
#add the English names of countries in nuts dataframe
final_spatial_data <- data_nuts_merged %>% 
                        inner_join(country_codes_names,
                                   by = "CNTR_CODE")
```






