---
title: "Getting Started with giscoR Maps"
author: "Shefali C."
date: "2023-08-21"
output: 
      html_document:
        toc: TRUE
        toc_depth: 2
        theme: readable
        highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', fig.width = 5, fig.height = 5)
```

### Introduction

This notebook contains an introduction to the **giscoR package**. Most of the code is available in bits and pieces at the following links:  

1. [giscoR website](https://ropengov.github.io/giscoR/)  
2. [rdrr.io](https://rdrr.io/cran/giscoR/man/)  

### What is "gisco"?  

***gisco*** stands for **Geographic Information System of the COmission**. This department is a part of **Eurostat** & is responsible for creating statistical and thematic maps for the European Union, manage a database of geographic information etc. Read more about it [here](https://ec.europa.eu/eurostat/web/gisco).  

### But what is "Eurostat"?  

**Eurostat** is the statistical office of the European Union. They are incharge of providing the EU with statistical information and data on Europe. Read more about it [here](https://ec.europa.eu/eurostat/about-us/who-we-are). 


```{r load-libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(giscoR)
library(sf)
library(kableExtra)
```

### gisco_db  

**`gisco_db`** gives a list of all datasets available with **giscoR** package. It contains countries data, coastline data, NUTS levels for years upto 2020 or 2021 depending on the attribute you are dealing with.  
For e.g., at the time of this writing, countries data is available upto year 2020 and NUTS data is available (ie. updated ) upto 2021.  

10 variables are available:  

<div style="float: left; width: 50%;">
  - id_giscoR  
  - year  
  - epsg  
  - resolution  
  - spatialtype
</div>

<div style="float: right; width: 50%;">
  - nuts_level  
  - level  
  - ext  
  - api_file  
  - api_entry
</div>


A glimpse of what gisco_db contains...  

Note: This table doesn't show the entire dataframe content. I've selected only 1 row per group, ie. 1 row for each **`id_giscoR`** type. 

```{r gisco-db2, echo=TRUE}
  gisco_db %>% 
  group_by(id_giscoR) %>% 
  select(id_giscoR, year, epsg, resolution) %>% 
  slice_head(n = 1) %>% 
  kbl() %>% 
  kable_paper("hover", full_width = F)
```


### gisco_countries data  

**`gisco_countries`**- It returns an sf object contains information and polygon shapefile of all countries returned by GISCO.

```{r brazil-map}
countries <- gisco_countries

#extract data for Brazil
brazil <- subset(countries, ISO3_CODE == "BRA")

#map of brazil
ggplot(data = brazil)+
  geom_sf(color = "goldenrod3", fill = "springgreen4", linewidth = 1.5)+
  theme_minimal()
```

### Using different resolutions

Some parameters of **`gisco_get_countries()`** and the values they accept are listed below:  

1. **spatialtype-** refers to the type of geometry to be returned.  
      i)  **"BN"**- boundaries; LINESTRING object.  
      ii) **"COASTL"**- coastlines; LINESTRING object.  
      iii) **"INLAND"**- inland boundaries; LINESTRING object.  
      iv) **"LB"**- labels; POINT object.  
      v) **"RG"**- shape of countries/regions; MULTIPLOYGON/POLYGON object.  
<br>
2. **resolution-** refers to the resolution of the geospatial data. 1 unit on the map represents what area on ground.  
      i)   **"60"**: 1:60million  
      ii)  **"20"**: 1:20million  
      iii) **"10"**: 1:10million  
      iv)  **"03"**: 1:3million  
      v)   **"01"**: 1:1million 
      
<br>

3. **epsg-** refers to the projection of the map. One of the following projections can be supplied to this parameter.  

    - **"4258"**: ETRS89  
    - **"4326"**: WGS84  
    - **"3035"**: ETRS89 / ETRS-LAEA  
    - **"3857"**: Pseudo-Mercator 


<br>

Read more about **`gisco_get_countries()`** [here](https://rdrr.io/cran/giscoR/man/gisco_get.html).  
You may go through the function code on github to understand how the function works. [Github](https://github.com/rOpenGov/giscoR/blob/main/R/gisco_get_countries.R)  


Country being mapped below- Germany

```{r germany-data}
#dat for Germany at different resolutions
deu_res60 <- gisco_get_countries(country = "DEU", resolution = "60") %>% 
                  mutate(res = "60m")

#resolution 20m
deu_res20 <- gisco_get_countries(country = "DEU", resolution = "20") %>% 
                  mutate(res = "20m")

#resolution 3m
deu_res3 <- gisco_get_countries(country = "DEU", res = "03") %>% 
                  mutate(res = "3m")

#resolution 1m
deu_res1 <- gisco_get_countries(country = "DEU", resolution = "01") %>% 
                  mutate(res = "1m")

#combined data for all resolutions
deu_all <- bind_rows(deu_res1, deu_res3, deu_res20, deu_res60)
```

```{r germany-map}
ggplot(data = deu_all)+
  geom_sf(fill = "gold")+
  facet_wrap(~res)+
  theme_minimal()+
  theme(
    axis.text = element_text(size = 6, face = "bold", color = "grey90"),
    plot.background = element_rect(fill = "black"),
    panel.background = element_rect(fill = "black"),
    panel.grid = element_blank(),
    strip.text = element_text(color = "grey90")
  )
```

### Maps of multiple countries

```{r ind-pak-chin-data}

india <- gisco_get_countries(country = "IND", resolution = "01")
pak <- gisco_get_countries(country = "PAK", resolution = "01")
china <- gisco_get_countries(country = "CHN", resolution = "01")

ind_pak_chn <- bind_rows(india, pak, china)

```

Well, the India-Pakistan-China map is quite weird. We definitely have disputed land area between India and Pakistan (Pakistan-occuped-Kashmir) and between India-China (Askai-Chin). But it seems like rest of the world doesn't identify Ladhakh and major part of Jammu-Kashmir to be part of India. It has been completely omitted. 

```{r ind-pak-chn-map}
ggplot(ind_pak_chn)+
  geom_sf(fill = c("IND" = "lightblue", "PAK" = "lightgreen", "CHN" = "red"))+
  theme_minimal()+
  theme(axis.text = element_text(size = 6, face = "bold", color = "black"))

```

### World Coastline

Following **`spatialtypes`** can be used to fetch different kinds of data from **`gisco_get_countries()`** function.  



```{r world-coastlines-data}

all_coast <- gisco_get_countries(
  spatialtype = "COASTL",
  #set projection to WGS 84, used in GPS
  epsg = "4326"
)

```

```{r world-coastlines-map}
ggplot(all_coast)+
  geom_sf(color = "midnightblue")+
  theme_minimal()+
  theme(#plot.background = element_rect(fill = "#F6E1B9"),
        panel.background = element_rect(fill = "#F6E1B9")
        )

```

### African Coastline with city labels  

```{r africa-coast-data}

#get city labels
africa_labs <- giscoR::gisco_get_countries(
  spatialtype = "LB",
  #Web Mercartor projection, used by Google Maps, OpenStreetMaps etc.
  epsg = "3857",
  region = "Africa"
)

#get african coastline LINE shapefile
africa_coast <- giscoR::gisco_get_countries(
  spatialtype = "COASTL",
  epsg = "3857"
)

```  

The **`st_bbox`** function of **sf** package is used to fetch the vertical (latitudes) and horizontal (longitudes) extent of a region. It returns 4 values: xmax, xmin, ymax, ymin.  

These values can be used zoom into any region on the world map. Or better say, crop the world map to focus on the desired region.  



```{r africa-bounding-box}

#returns the extent of Africa
st_bbox(africa_labs)

#This basically zooms in the otherwise world map to focus on the Africa.
africa_bbox <- sf::st_bbox(africa_labs)

```

```{r africa-coast-map}

#Map of African coast
ggplot(africa_coast)+
  #aesthetic "color" used & not "fill" because coastlines are line shapefiles.
  geom_sf(colour = "deepskyblue4", linewidth = 2)+
  geom_sf(data = africa_labs, fill = "springgreen4", colour = "darkgoldenrod1", shape = 21, size = 3)+
  coord_sf(xlim = africa_bbox[c("xmin", "xmax")],
           ylim = africa_bbox[c("ymin", "ymax")])+
  theme_minimal()

```

### European NUTS levels  

NUTS stands for "Nomenclature des Unités territoriales statistiques" in French. In english, it translates to " Nomenclature of territorial units for statistics".  

The economic territory of EU has been divided into regions at 3 levels- NUTS 1, NUTS 2 and NUTS 3. Above NUTS 1, we have the national level.  
(It can be considered analogous to a division like- Country-->States-->Cities-->Wards.) 

```{r italy-nuts1-data}
#get NUTS level 1 region names for Italy
italy_nuts1 <- giscoR::gisco_get_nuts(country = "Italy", nuts_level = 1)

```

```{r italy-nuts1-map}
#labelled map
ggplot(italy_nuts1)+
  geom_sf(aes(fill = NAME_LATN), color = "black")+
  geom_sf_text(aes(label = NAME_LATN), size = 4)+
  theme_minimal()+
  theme(axis.title = element_blank())
```

### Oceania Coastline

```{r oceania-coastline-map}
#World coastlines
world_coasts <- gisco_coastallines


#Zoom in on Oceania coastlines
ggplot(world_coasts)+
  geom_sf(color = "cyan", fill = "cyan", alpha = 0.2)+
  #zoom on Oceania part
  coord_sf(xlim = c(96, 179),
           ylim = c(-51, 11)
           )+
  theme_minimal()+
  theme(
    plot.background = element_rect(fill = "black", color = "black"),
    panel.grid = element_blank(),
    axis.text = element_text(color = "grey90")
  )

```

### Greece Airports

```{r greece-airport-data}

greece <- gisco_get_countries(country = "EL", res = "1")
#airports of greece
greece_airport <- gisco_get_airports(country = "EL")
#transform the CRS
greece_airport <- st_transform(greece_airport, st_crs(greece))

```

```{r greece-airport-map}
#map
ggplot(greece)+
  geom_sf(fill = "grey80")+
  geom_sf(data = greece_airport, color = "blue")+
  labs(title = "Airports of Greece",
       caption = gisco_attributions())+
  theme_minimal()+
  theme(plot.caption = element_text(hjust = 0))
```

### World Seaports

```{r world-sea-ports}

world_ports <- gisco_get_ports(year = "2013")
world_coasts <- gisco_get_coastallines(year = "2013")

#For equal Earth projection
world_coasts <- st_transform(world_coasts, 8857)
world_ports <- st_transform(world_ports, st_crs(world_coasts))

```

```{r world-seaports-map}

ggplot(data = world_coasts)+
  geom_sf(fill = "#F6E1B9", color = "#0978AB")+
  geom_sf(data = world_ports, fill = "red", shape = 21)+
  #theme_void()+
  theme(
    panel.background = element_rect(fill = "#C6ECFF"),
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(face = "italic", hjust = 0.5)
  ) +
  labs(
    title = "Ports Worldwide", subtitle = "Year 2013",
    caption = "(c) European Union, 1995 - today"
  )
```

### Carribean Coastlines 

```{r carribean-coast}
coast <- gisco_get_coastallines()

ggplot(coast) +
  geom_sf(color = "#1278AB", fill = "#FDFBEA") +
  # Zoom on Caribe
  coord_sf(
    xlim = c(-99, -49),
    ylim = c(4, 30)
  ) +
  theme_minimal() +
  theme(panel.background = element_rect(fill = "#C7E7FB", color = "black"))
```

### Using gisco_get_grid()  

Application of  **`gisco_get_grid()`** function deserves its own notebook. You can go through [this](add link to your notebook) to see its usage.  
A short note on the parameter value it accepts is as follows:  


- **resolution** accepts values- "1", "2", "5", "10", "20", "50", "100". `resolution = "100"` results in file-size 428Kb & `resolution = "1"` results in file-size of about 1.7Gb.  

- **spatialtype** can be either "REGION" or "POINT".  

- The projection of the files returned is **EPSG: 3035**.  

### Location of heathcare services in Europe

- Function used: **`gisco_get_healthcare()`**.  
- Projection: **EPSG-4326**  
- Value returned- POINT object.  
- The dataframe contains 26 variables with info about hospital names, location, lat-long values, kind of ownership (private/public), email, phone number, city, street etc. 
<br>
Further reading: [rdrr.io](https://rdrr.io/cran/giscoR/man/gisco_get_healthcare.html)

```{r belgium-health-data}

belgium_healthcare <- gisco_get_healthcare(country = "Belgium")
#replace blank in kind of ownership column to "unknown"
belgium_healthcare[belgium_healthcare$public_private == "",]$public_private <- "unknown"

#get NUTS level 2 region names of belgium
belgium_nuts2 <- gisco_get_nuts(country = "Belgium", nuts_level = 2)
```

```{r belgium-health-map}
ggplot(data = belgium_nuts2)+
  geom_sf(fill = "white", color = "black", linewidth = 0.5)+
  geom_sf(data = belgium_healthcare, aes(color = public_private),
          alpha = 0.6, size = 2)+
  theme_bw()+
  labs(title= "Healthcare in Belgium",
       caption = paste(gisco_attributions()),
       subtitle = "NUTS 2",
       fill = "type")+
  scale_color_manual(name = "Type of ownership", values = hcl.colors(n = 3, palette = "Berlin"))+
  theme_minimal()+
  theme(
    plot.title = element_text(size = 11),
    plot.subtitle = element_text(size = 9),
    plot.caption = element_text(size = 8, hjust = 0),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 8, face = "bold"),
    legend.text = element_text(size = 7)
  )

```

### Local Urban Areas of Europe

- This data can be obtained using functions- **`gisco_get_lau()`** & **`gisco_get_communes()`**.  
- 

```{r ireland-urban-areas-data}

#get labels/names for the urban areas of Ireland
#ireland_uareas <- gisco_get_lau(country = "Ireland")


```























